Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
update camera-ready version after fixing all the spelling and grammatical errors

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of observations generated by molecular, physiological or psychological processes to make predictions about the outcome of these processes arises in many domains of clinical informatics. In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog. Specifically, we consider prediction of the motivational interview success (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem. We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews. Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate that the proposed method can be used for real-time monitoring of progression of clinical interviews and more efficient identification of effective provider communication strategies, which in turn can significantly decrease the effort required to develop behavioral interventions and increase their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog). Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations. Predicting the outcome of those sequences (e.g. physiological or behavioral response) can also be viewed as a sequence classification problem.

In general, sequence classification methods fall into one of three major classes: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at classification decision. The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
sequence.

Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by timestamps. The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then classifies new sequences by applying the created models. While MCs and HMMs can capture first- and second-order dependencies between adjacent observations in a sequence, learning higher-order dependencies with these models requires prohibitively large amounts of data. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short- and long-term dependencies and were shown to be effective at modeling different types of sequential data \cite{lipton2015critical}. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which successfully addressed the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and handwriting recognition\cite{nion2013handwritten} to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit (GRU)\cite{chung2014empirical}, was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of clinical dialog. Specifically, we focus on predicting the success (i.e. eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded PPC exchanges during those interviews. Childhood obesity is a serious public health concern in the United States. Recent estimates indicate that approximately one-third (31.8\%) of U.S. children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e. eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous PPC exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.

@@

Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g. simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address class imbalance. We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes. Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes. Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.

@@

\textbf {Behavior code embeddings}. Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}. Embedding is a representation of an object in low-dimensional space using a real-valued vector. In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as representation of behavior codes as input to these RNNs. Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left part of Figure~\ref{fig:code_embedding}. The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e. OQECML+ being more purple indicates that it is more similar to CQECML+). The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.

@@

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their manually annotation of experimental dataset with MYSCOPE codes.

>>>>

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided an equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of observations generated by molecular, physiological or psychological processes to make predictions about the outcome of these processes arises in many domains of clinical informatics. In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of the clinical dialog. Specifically, we consider prediction of the motivational interview success (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem. We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews. Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate that the proposed method can be used for real-time monitoring of progression of clinical interviews and more efficient identification of effective provider communication strategies, which in turn can significantly decrease the effort required to develop behavioral interventions and increase their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances in the clinical dialog). Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than a classification of independent observations. Predicting the outcome of those sequences (e.g. physiological or behavioral response) can also be viewed as a sequence classification problem.

In general, sequence classification methods fall into one of three major classes: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at classification decision. The methods in this class have had limited success since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
sequence.

Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by timestamps. The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then, classifies new sequences by applying the created models. While MCs and HMMs can capture first- and second-order dependencies between adjacent observations in a sequence, learning higher-order dependencies with these models requires prohibitively large amounts of data. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short- and long-term dependencies and were shown to be effective at modeling different types of sequential data \cite{lipton2015critical}. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which successfully addressed the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and handwriting recognition\cite{nion2013handwritten} to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from the previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit (GRU)\cite{chung2014empirical}, was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of the clinical dialog. Specifically, we focus on predicting the success (i.e. eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded PPC exchanges during those interviews. Childhood obesity is a serious public health concern in the United States. Recent estimates indicate that approximately one-third (31.8\%) of U.S. children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e. eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous PPC exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored the applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.

@@

Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g. simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address the class imbalance. We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes. Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes. Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.

@@

\textbf {Behavior code embeddings}. Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}. Embedding is a representation of an object in low-dimensional space using a real-valued vector. In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as a representation of behavior codes as input to these RNNs. Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left part of Figure~\ref{fig:code_embedding}. The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e. OQECML+ being more purple indicates that it is more similar to CQECML+). The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.

@@

\section*{Acknowledgments}
This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their manual annotation of experimental dataset with MYSCOPE codes.
